---
title: "lmer Model"
output: word_document
---

This is the code for modeling our data using lmer function. The model we are using is:

$$A_i = b_i + \beta_T + b_T_i + \mu + \epsilon_i$$
$$B_i = b_i + \beta_T + b_T_i + \mu + \epsilon_i$$
$$N_i = b_i ~~~~~~~~~~~~~~~~~~+ \mu + \epsilon_i$$


* $A_i$, $B_i$, $N_i$ are the mythelation for side A, B, and N respectively
* $\beta_T$ = effect of tumor tissue (fixed)
* $\mu$ = grand mean
* $b_i \sim N(0,\sigma^2)$
* $b_T_i \sim N(0,\sigma^2_T)$
* $\epsilon_i \sim N(0,\sigma^2_y)$

The codes below only use patient and tumor tissue levels.

Please make sure the annotation file is saved in the folder where all the other data are stored. 

* fit1: X1 ~ (1|patient) + tInd + (tInd-1|patient)
* fit2: X1 ~ tInd + (tInd|patient)
* fit3: X1 ~ (tInd|patient)

``` {r, echo=FALSE, warning=FALSE, message=FALSE}

### install the following package
# install.packages("lme4")

library(lme4)

# here set the working directory that points to the data folder
# e.g. the folder with annotated data saved as "myFA.Rdata"
# please just comment out the directories already here, so each
# user can uncomment the setwd for them as we move code around

# Kevin's working directory
# setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")

# Yanlin's working directory
setwd("D:/DataPlus2017/Data")

# Set your directory here


### LOAD DATA

# load fully annotated data (saved from LoadDataAndQC.R)
load("myFA.Rdata")

# extract CpG site 1 to start
firstSite <- head(FullAnnotation, 1)

# convert first site to table, using only H E K* W C J (tumor+normal)
# 18 rows, with columns for patient, tumor/normal
# index in order A,B,N
indices <- c(11, 12, 13, 17, 18, 16, 20, 21, 19, 25, 26, 24, 33, 34, 32, 
             39, 31, 15)
patientLabel <- c(rep("H", 3),rep("E", 3),rep("K*", 3),rep("W", 3),rep("C", 3),
                  rep("J", 3))
tissueLabel <- rep(c("T", "T", "N"), 6)
sideLabel <- rep(c("A", "B", ""), 6)

# here we use all samples
indices <- c(9:13, 15:39,46,47,57,58,72:75)
patientLabel <- substr(colnames(firstSite[indices]),1,1)
patientLabel[10:12] <- "K*"
sideLabel <- substr(colnames(firstSite[indices]),2,2)
tissueLabel <- sideLabel
tissueLabel[tissueLabel %in% c("A","B")] <- "T"   #replace A and B with T
tissueIndicator <- 1*(tissueLabel=="T")

firstData <- data.frame(beta=t(firstSite[indices]), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tissueIndicator)


### FIT WITH LMER

fit1 <- lmer(X1 ~ (1|patient) + tInd + (tInd-1|patient), firstData)
fit2 <- lmer(X1 ~ tInd + (tInd|patient), firstData)
fit3 <- lmer(X1 ~ (tInd|patient), firstData)

patLabs <- c("C", "D","E","F","H","J","K","K*","M","O","P","S","T","U","W","X")

```


#### Summary for fit1 

``` {r,echo=FALSE}
summary(fit1)

```


#### Summary for fit2 

``` {r,echo=FALSE}
summary(fit2)

```


#### Summary for fit3 

``` {r,echo=FALSE}
summary(fit3)

```



#### Plots for Estimates

``` {r, echo=FALSE}

# barplot(fixef(fit1), main="Fixed Effects", xlab="Effect", ylab="Intercept Estimate", names.arg=c("mu","betaT"))
# barplot(ranef(fit1)$patient$'(Intercept)', main="Patient Random Effect", xlab="Patient", ylab=expression(paste("Intercept Estimate (",b[i],")")), cex.names = 0.7, las=2)
# barplot(ranef(fit1)$patient$tInd, main="Patient,Tumor Slope Random Effect", xlab="Patient", ylab=expression(paste("Slope Estimate (",b[iT],")")), cex.names = 0.7, las=2)

barplot(fixef(fit1), main="Fixed Effects (fit1)", xlab="Effect", ylab="Intercept Estimate", names.arg=c("mu","betaT"), ylim=c(-0.2,0.2))
barplot(fixef(fit2), main="Fixed Effects (fit2)", xlab="Effect", ylab="Intercept Estimate", names.arg=c("mu","betaT"), ylim=c(-0.2,0.2))
barplot(ranef(fit3)$patient$'(Intercept)', main="Patient Random Effect (fit3)", xlab="Patient", ylab=expression(paste("Intercept Estimate (",b[i],")")), names.arg = patLabs, cex.names = 0.7, ylim=c(-0.2,0.2))
barplot(ranef(fit2)$patient$'(Intercept)', main="Patient Random Effect (fit2)", xlab="Patient", ylab=expression(paste("Intercept Estimate (",b[i],")")), names.arg = patLabs, cex.names = 0.7, ylim=c(-0.2,0.2))
barplot(ranef(fit3)$patient$tInd, main="Patient,Tumor Slope Random Effect (fit3)", xlab="Patient", ylab=expression(paste("Slope Estimate (",b[iT],")")), names.arg = patLabs, cex.names = 0.7, ylim=c(-0.2,0.2))
barplot(ranef(fit2)$patient$tInd, main="Patient,Tumor Slope Random Effect (fit2)", xlab="Patient", ylab=expression(paste("Slope Estimate (",b[iT],")")), names.arg = patLabs, cex.names = 0.7, ylim=c(-0.2,0.2))

barplot(ranef(fit1)$patient$'(Intercept)'+ranef(fit1)$patient$tInd, main="Both Random Effects (fit1)", xlab="Patient", ylab=expression(paste("Both Estimates (",b[i]+b[iT],")")), names.arg = patLabs, cex.names = 0.7, ylim=c(-0.2,0.2))
barplot(ranef(fit2)$patient$'(Intercept)'+ranef(fit2)$patient$tInd, main="Both Random Effects (fit2)", xlab="Patient", ylab=expression(paste("Both Estimates (",b[i]+b[iT],")")), names.arg = patLabs, cex.names = 0.7, ylim=c(-0.2,0.2))

#barplot(fixef(fit1), main="Patient Fixed Effects", xlab="Patient", ylab=expression(paste("Intercept Estimate (",beta[k],")")), names.arg=c("C", "D","E","F","H","J","K","K*","M","O","P","S","T","U","W","X"), cex.names=0.7)
#barplot(fixef(fit1)[-17]+fixef(fit1)[1], main="Patient Fixed Effects", xlab="Patient", ylab=expression(paste("Intercept Estimate (",beta[k],")")), names.arg=c("C","D","E","F","H","J","K","K*","M","O","P","S","T","U","W","X"), cex.names=0.65)


```